The need for accuracy and smoothness in numerical simulations

### 2024-09-05 slides/ostrava/cckm-ppam-2024.tex
Slides for PPAM-2024 approaching final shape. Restructured slide folder to accommodate figures shared between talks

\graphicspath{{../../report/fig/}}

@@

\frame{\frametitle{Warning: Disturbing pictures to follow}

  We did two sets of simulations of a molecule in water
  \vspace{1cm}
  \begin{enumerate} \setlength\itemsep{1em}
    \item fixed real time
    \item two different force-fields: \texttt{charmm36}, \texttt{oopsaal}
    \item identical initial conditions
   \end{enumerate}
  \vspace{1cm}
  and tracked the energy as a function of the number of steps.

@@

\frame{\frametitle{Warning: Disturbing pictures below}

 \begin{figure}
    \includegraphics[width=8cm]{oplsaaltol12.pdf}
  \end{figure}

  \begin{figure}
    \includegraphics[width=8cm]{charmm36tol12.pdf}
  \end{figure}

@@

@@

  \begin{block}{Key questions}
    Does it matter
    \begin{enumerate}
    \item if the force-field 
      \begin{equation*}
        \bm{q} \rightarrow \bm{f}(\bm{q})
      \end{equation*}
      is smooth or not?
    \item if the nonlinear constraint equation
      \begin{equation}
        \bm{g}(\bm{q}_{n+1}(\bm{\lambda})) = \bm{0}
      \end{equation}
      is solved accurately with respect to $\bm{\lambda}$ or not?
    \end{enumerate}
  \end{block}
}

@@

  If one of the following is true:
  \vspace{0.5cm}
  \begin{enumerate}
  \item the central functions are not smooth enough
  \item the central equations are not solved accurately enough
  \end{enumerate}
  \vspace{0.5cm}
  then we cannot:
  \vspace{0.5cm}
  \begin{enumerate}
  \item estimate the modelling error
  \item validate the model against the physical reality
  \end{enumerate}
  \vspace{1cm}
}

@@

  We validate our models by demonstrating that
  \begin{equation}
    P - T \approx 0
  \end{equation}
  is a good approximation, but
  \begin{equation}
    P - T \approx P - \hat{A}_h
  \end{equation}
  is not necessarily a good approximation. We have

  \begin{equation}
    P - T = (P - \hat{A}_h) - (T - A_h) - (A_h - \hat{A}_h)
  \end{equation}
  so we need to assert that
  \begin{align}
    |T- A_h| & \ll |P - \hat{A}_h| \\
    |A_h - \hat{A}_h| & \ll |P - \hat{A}_h|
  \end{align}
}

@@

  It is frequently possible to simultanously
  \begin{enumerate}
  \item assert that the computational error
    \begin{equation*}
      A_h - \hat{A}_h
    \end{equation*}
    is irrelevant, and
  \item estimate the discretization error
    \begin{equation*}
      T_h - A_h
    \end{equation*}
    accurately
  \end{enumerate}
  The key is the existence of an asymptotic error expansion
  \begin{equation}
    E_h = T - A_h = \alpha h^p + \beta h^q + O(h^r), \quad h \rightarrow 0_+
  \end{equation}
  where $0 < p < q < r$.
}

@@

Define 
\begin{equation}
  R_h := \frac{A_h - A_{2h}}{2^p - 1}, \quad F_h:= \frac{A_{2h} - A_{4h}}{A_h - A_{2h}} 
\end{equation}
If 
\begin{equation}
  E_h = T - A_h = \alpha h^p + \beta h^q + O(h^r), \quad h \rightarrow 0_+
\end{equation}
then
\begin{equation}
  \frac{E_h - R_h}{h^q} \rightarrow \text{constant}
\end{equation}
and
\begin{equation}
  \frac{2^p - F_h}{h^{q-p}} \rightarrow \text{constant}
\end{equation}
which implies
\begin{equation}
  \log | 2^p - F_h| \approx \log|\text{constant}| + (q-p) \log h
\end{equation}
}

@@

  \begin{figure}[ht]
        \begin{minipage}[b]{0.45\linewidth}
            \centering
            \includegraphics[width=\textwidth]{rint_mwe1a.pdf}
            \caption{The evolution of $F_h$}
            \label{fig:a}
        \end{minipage}
        \hspace{0.5cm}
        \begin{minipage}[b]{0.45\linewidth}
            \centering
            \includegraphics[width=\textwidth]{rint_mwe1b.pdf}
            \caption{The accuracy of $R_h$}
            \label{fig:b}
        \end{minipage}
    \end{figure}
    Strictly speaking we are not observing $F_h$ and $R_h$
  \begin{center}
    we are observing $\hat{F}_h$ and $\hat{R}_h$
  \end{center}
  The discrepancy is controlled by the computational error
  $$ A_h - \hat{A}_h$$
}

@@

  GROMACS simulation of hen egg white lysozyme in water:
  \vspace{0.25cm}
  \begin{enumerate} \setlength\itemsep{1em}
  \item $T$: total energy of the simulation at the end
  \item $A_h$: the approximation of $T$ computed using SHAKE
  \item $\hat{A}_h$: the value of $A_h$ returned by the computer
  \end{enumerate}
\vspace{0.5cm}
Temporal matters:
\vspace{0.25cm}
  \begin{enumerate}
  \item The length of the simulations was $10^{-12}$ s ($1$ ps)
  \item A common step size in MD is $10^{-15} s$ ($1$ fs)
  \end{enumerate}

  \begin{center}
    So, a mere 1000 steps! What could possibly go wrong?
  \end{center}
}

@@

  Deriving an AEX is an exercise in Taylor expansions. If
  \begin{center}
    our functions are not many times differentiable
  \end{center}
  then the foundation crumbles.
  \vspace{1cm}
    \newline
  Every AEX refers to the exact value of the $A_h$. If
  $$\hat{A}_h \approx A_h$$
  is not a good approximation, then
  \begin{center}
    $\hat{A}_h$ will not behave nicely
  \end{center}
}

@@

\frame{\frametitle{Computing the optimal range of the a howitzer}

 \begin{figure}
    \includegraphics[width=10cm]{maxrange_rk1_tol18.pdf} \caption{The evolution of Richardson's fraction for 3 different drag coeffients, Euler's explicit method and sufficiently accurate event location.}
  \end{figure}

@@

\frame{\frametitle{Computing the optimal range of the a howitzer}

 \begin{figure}
    \includegraphics[width=10cm]{maxrange_rk1_tol11.pdf} \caption{The evolution of Richardson's fraction for 3 different drag coeffients, Euler's explicit method and inaccurate event location.}
  \end{figure}

@@

\frame{\frametitle{Modelling ions}

\frame{\frametitle{Molecular dynamics: Truncation of force-fields}

  \begin{itemize}
  \item Simple simulation with 4 ions
  \item Each ion is attached to the origin with a spring
  \item The electrostatic forcefields are trunctated.
  \end{itemize}
}

@@

  We use every trick in the book to
  \begin{enumerate}
  \item reduce time-to-solution
  \item increase the parallel efficicency
  \item reduce energy-to-solution
  \end{enumerate}
  \vspace{1cm}
  We should not forget to ask the question:
  \begin{center}
    Did we just compromise our ability to validate our models against the real world?
  \end{center}
}

>>>>

\graphicspath{{../fig/}{../../report/fig/}}

@@

\frame{\frametitle{A simple experiment}

  We did two sets of simulations of a molecule in water
  \vspace{1cm}
  \begin{enumerate} \setlength\itemsep{1em}
    \item fixed real time
    \item two different force-fields: \texttt{charmm36}, \texttt{oopsaal}
    \item identical initial conditions
    \item ran the constraint solver to stagnation
   \end{enumerate}
  \vspace{1cm}
  and tracked the final energy as a function of the number of steps.

@@

\frame{% \frametitle{Warning: Disturbing pictures below}

 \begin{figure}
    \includegraphics[width=10.5cm]{oplsaaltol12.pdf}
  \end{figure}

  \begin{figure}
    \includegraphics[width=10.5cm]{charmm36tol12.pdf}
  \end{figure}

@@

\frame{\frametitle{Molecular dynamics with constraints}

@@

  \begin{block}{Key questions}
    Does it matter\vspace{0.5cm}
    \begin{enumerate}\setlength\itemsep{1.5em}
    \item if the force-field 
      \begin{equation*}
        \bm{q} \rightarrow \bm{f}(\bm{q})
      \end{equation*}
      is smooth or not?
    \item if the nonlinear constraint equation
      \begin{equation*}
        \bm{g}(\bm{q}_{n+1}(\bm{\lambda})) = \bm{0}
      \end{equation*}
      is solved accurately with respect to $\bm{\lambda}$ or not?
    \end{enumerate}
  \end{block}
}

@@

  If one of the following is true:
  \vspace{0.5cm}
  \begin{enumerate}
  \item the central functions are not smooth enough
  \item the central equations are not solved accurately enough
  \end{enumerate}
  \vspace{0.5cm}
  then we may lose the ability to
  \vspace{0.5cm}
  \begin{enumerate}
  \item assert that rounding errors are irrelevant
  \item estimate the discretization error
  \item compare the model against the physical reality
  \end{enumerate}
  \vspace{1cm}
}

@@

  We validate our models by demonstrating that
  \begin{equation}
    P - T \approx 0
  \end{equation}
  is a good approximation, but
  \begin{equation}
    P - T \approx P - \hat{A}_h
  \end{equation}
  is not necessarily a good approximation. We have
  \begin{equation}
    P - T = (P - \hat{A}_h) - (T - A_h) - (A_h - \hat{A}_h)
  \end{equation}
  so we need to assert that
  \begin{align}
    |T- A_h| & \ll |P - \hat{A}_h| \\
    |A_h - \hat{A}_h| & \ll |P - \hat{A}_h|
  \end{align}
}

@@

  It is frequently possible to simultanously
   \vspace{0.5cm}
  \begin{enumerate} \setlength\itemsep{1.5em}
  \item assert that the computational error
    \begin{equation*}
      A_h - \hat{A}_h
    \end{equation*}
    is irrelevant, and
  \item estimate the discretization error
    \begin{equation*}
      T_h - A_h
    \end{equation*}
    accurately
  \end{enumerate}
}

  The key is to have an asymptotic error expansion (AEX)
  \begin{equation}
    E_h = T - A_h = \alpha h^p + \beta h^q + O(h^r), \quad h \rightarrow 0_+
  \end{equation}
  where
  \begin{equation}
    0 < p < q < r
  \end{equation}
  are not necessarily integers.
}

@@

Define 
\begin{equation}
  R_h := \frac{A_h - A_{2h}}{2^p - 1}, \quad F_h:= \frac{A_{2h} - A_{4h}}{A_h - A_{2h}} 
\end{equation}
Then
\begin{equation}
  \frac{E_h - R_h}{h^q} \rightarrow \text{constant}
\end{equation}
and
\begin{equation}
  \frac{2^p - F_h}{h^{q-p}} \rightarrow \text{constant}
\end{equation}
which implies
\begin{equation}
  \log | 2^p - F_h| \approx \log|\text{constant}| + (q-p) \log h
\end{equation}
}

@@

  \begin{figure}[ht]
        \begin{minipage}[b]{0.45\linewidth}
            \centering
            \includegraphics[width=\textwidth]{rint_mwe1a.pdf}
            \caption{The evolution of $F_h$}
            \label{fig:a}
        \end{minipage}
        \hspace{0.5cm}
        \begin{minipage}[b]{0.45\linewidth}
            \centering
            \includegraphics[width=\textwidth]{rint_mwe1b.pdf}
            \caption{The accuracy of $R_h$}
            \label{fig:b}
        \end{minipage}
    \end{figure}
    Strictly speaking we are not observing $F_h$ and $R_h$
  \begin{center}
    we are observing $\hat{F}_h$ and $\hat{R}_h$
  \end{center}
  The difference is controlled by the computational error
  \begin{equation}
    A_h - \hat{A}_h
  \end{equation}
}

@@

  GROMACS simulation of hen egg white lysozyme in water:
  \vspace{0.25cm}
  \begin{enumerate} \setlength\itemsep{1em}
  \item $T$: total energy of the simulation at the end
  \item $A_h$: the approximation of $T$ computed using SHAKE
  \item $\hat{A}_h$: the value of $A_h$ returned by the computer
  \end{enumerate}
\vspace{0.5cm}
Temporal matters:
\vspace{0.25cm}
  \begin{enumerate}
  \item The length of the simulations was $10^{-12}$ s ($1$ ps)
  \item A common step size in MD is $10^{-15} s$ ($1$ fs) or $n=1000$ steps
  \item $n \in \{250,500,1000:100:2000,3000:1000:16000\}$
  \end{enumerate}

  \begin{center}
   What could possibly go wrong?
  \end{center}
}

@@

  \begin{enumerate} \setlength\itemsep{1.5em}
  \item Every AEX refers to the exact value of the $A_h$. If
    $$\hat{A}_h \approx A_h$$
    is not a good approximation, then
    \vspace{0.25cm}
    \begin{center}
      $\hat{A}_h$ will not behave nicely
    \end{center}
    \vspace{0.25cm}
    and we cannot identify the asymptotic range.
  \item Deriving an AEX is an exercise in Taylor expansions. If
    \vspace{0.25cm}
    \begin{center}
      our functions are not many times differentiable
    \end{center}
    \vspace{0.25cm}
    then the foundation crumbles.
  \end{enumerate}
}

@@

\frame{\frametitle{Computing the optimal range of the a howitzer: Success}

 \begin{figure}
    \includegraphics[width=10cm]{maxrange_rk1_tol18.pdf} \caption{The evolution of $F_h$ for 3 different drag coeffients, 1st order Runge-Kutta and sufficiently accurate event location.}
  \end{figure}

@@

\frame{\frametitle{Computing the optimal range of the a howitzer: Failure}

 \begin{figure}
    \includegraphics[width=10cm]{maxrange_rk2_tol18.pdf} \caption{The evolution of $F_h$ for 3 different drag coeffients, 2nd order Runge-Kutta and inaccurate event location.}
  \end{figure}

@@

\frame{\frametitle{Modelling ions: Setup}

  The total force on an ion is
  \begin{equation}
    \bm{F}(\bm{r_i}) =  - \alpha \sum_{j \not = i} \frac{1}{\|\bm{r}_i - \bm{r}_j\|^3}(\bm{r}_i - \bm{r}_j) - \beta \bm{r_i} - \gamma \bm{v}_i
  \end{equation}
  We measure the total kinetic energy $A_h$ at a fixed time as function of the stepsize $h_k = 2^{-k} h_0$. 
}

\frame{\frametitle{Modelling ions: Success}

   \begin{figure}
     \includegraphics[width=10cm]{iontrap_mwe1.pdf}
     \caption{The evolution of $F_h$ for Runge-Kutta methods of order $p \in \{1,2,3,4\}$ and smooth force-fields with infinite range.}
  \end{figure}
}

\frame{\frametitle{Modelling ions: Failure}
 \begin{figure}
    \includegraphics[width=10cm]{iontrap_mwe2.pdf} \caption{The evolution of $F_h$ for Runge-Kutta methods of order $p \in \{1,2,3,4\}$ and truncated force-fields with jump discontinuities.}
  \end{figure}
}

\frame{\frametitle{Modelling ions: Success}
 \begin{figure}
    \includegraphics[width=10cm]{iontrap_mwe4.pdf} \caption{The evolution of $F_h$ for Runge-Kutta methods of order $p \in \{1,2,3,4\}$ and smoothly truncated force-fields.}
  \end{figure}
}

@@

  We use every trick in the book to
  \vspace{0.5cm}
  \begin{enumerate} \setlength\itemsep{1.5em}
  \item reduce time-to-solution
  \item increase the parallel efficicency
  \item reduce energy-to-solution
  \end{enumerate}
  \vspace{1cm}
  We should not forget to ask the question:
  \begin{center}
    \begin{large}
      \textbf{Can we still validate our models against the real world?}
    \end{large}
  \end{center}
}
